\propertitle

### 2019-12-03 maxNt_v6.tex
update

\pdfoutput=1
%% Author: PGL  Porta Mana
%% Created: 2015-05-01T20:53:34+0200
%% Last-Updated: 2019-12-02T22:37:28+0100
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newif\ifarxiv
\arxivfalse
\ifarxiv\pdfmapfile{+classico.map}\fi
\newif\ifafour
\afourfalse% true = A4, false = A5
\newif\iftypodisclaim % typographical disclaim on the side
\typodisclaimtrue
\newcommand*{\memfontfamily}{zplx}
\newcommand*{\memfontpack}{newpxtext}
\documentclass[\ifafour a4paper,12pt,\else a5paper,10pt,\fi%extrafontsizes,%
onecolumn,oneside,article,%french,italian,german,swedish,latin,
british%
]{memoir}
\newcommand*{\updated}{\today}
\newcommand*{\firstdraft}{4 November 2015}
\newcommand*{\firstpublished}{***}
\newcommand*{\propertitle}{Maximum-entropy distributions for a 
  neuronal population\\ from subpopulation data {\large\color{myred}[draft]}%\\{\large ***}%
}
\newcommand*{\pdftitle}{\propertitle}
\newcommand*{\headtitle}{Maximum-entropy from sample data}
\newcommand*{\pdfauthor}{P.G.L.  Porta Mana, Y. Roudi, V. Rostami, E. Torre}
\newcommand*{\headauthor}{Porta Mana \etal}
\newcommand*{\reporthead}{}% Report number

@@

The limitations of approach here proposed can be appraised by examining the
assumptions and approximations on which it rests. Its assumptions and
approximations become clear within a full-fledged probabilistic inference
problem, which we sketch qualitatively.

The final assumption of the maximum-entropy method concerns the form of the
probability density $\pxx(\ynu)$ in the expressions above. It is taken to
be either an \enquote{entropic prior}
\citep{skillingetal1984,rodriguez1991,neumann2007}, which partially takes
into account the multiplicity of long-run frequencies, or a prior for a
model with sufficient statistics, which expresses the sufficiency of the
model under study \citep{portamana2017}. This assumption is crucial for
obtaining the maximum-entropy method with the Shannon entropy. A different
density would lead to alternative maximum-entropy methods, for example
based on the Burg entropy \citep{burg1975} if a Dirichlet density is used
\citep{jaynes1986d_r1996}.

In the study or use of the frequency distribution obtained with the
procedure here presented we must take into account two important points.

The first point is that there are many possible frequency distributions
which we believe, to different degrees, could be the true one that happened
during the recording. The one given by our procedure is simply the one with
the largest degree of belief, the mode of the belief distribution. The
space of possible frequency distributions has many dimensions, however --
thousands or tens of thousands. We must remember that belief distributions
in high dimensions have counter-intuitive properties. For example, the mode
or mean can have \emph{atypical} features when compared with the features
of most other points of the space. The mode and mean can also be very
different from each other.

The question, then, is \emph{which features of the maximum-entropy
  frequency distribution are typical of the majority of plausible frequency
  distributions?} We can only answer for sure by using the full-fledged
probability calculus. A more complete study (in preparation) with the first
data set reveals that most of the plausible frequency distributions have
three important features in common with the maximum-entropy one:
\begin{itemize}
\item all activities $\yA/N \gtrsim 5\%$ have practically zero frequencies;
\item there are two regions of activity levels with high frequencies,
  roughly separated by a trough of lower frequencies.
\item The frequencies of the region on the left ($\yA/N \lesssim 1.8\%$)
  are higher than those of the region on the right ($\yA/N \gtrsim 1.8\%$).
\end{itemize}
But there are also differences. For example, many plausible frequency
distributions have three or four modes instead of just two; these modes are
higher than those of the maximum-entropy distribution; and the bump of high
frequencies on the right is slightly shifted towards lower activities than
the corresponding maximum in the maximum-entropy distribution.

@@

>>>>

\pdfoutput=1
%% Author: PGL  Porta Mana
%% Created: 2015-05-01T20:53:34+0200
%% Last-Updated: 2019-12-03T09:50:35+0100
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newif\ifarxiv
\arxivfalse
\ifarxiv\pdfmapfile{+classico.map}\fi
\newif\ifafour
\afourfalse% true = A4, false = A5
\newif\iftypodisclaim % typographical disclaim on the side
\typodisclaimtrue
\newcommand*{\memfontfamily}{zplx}
\newcommand*{\memfontpack}{newpxtext}
\documentclass[\ifafour a4paper,12pt,\else a5paper,10pt,\fi%extrafontsizes,%
onecolumn,oneside,article,%french,italian,german,swedish,latin,
british%
]{memoir}
\newcommand*{\updated}{\today}
\newcommand*{\firstdraft}{4 November 2015}
\newcommand*{\firstpublished}{***}
\newcommand*{\propertitle}{Maximum-entropy distributions for a 
  neuronal population\\ from subpopulation data {\large\color{myred}[draft]}%\\{\large ***}%
}
\newcommand*{\pdftitle}{\propertitle}
\newcommand*{\headtitle}{Maximum-entropy from sample data}
\newcommand*{\pdfauthor}{P.G.L.  Porta Mana, Y. Roudi, V. Rostami, E. Torre}
\newcommand*{\headauthor}{Porta Mana \etal}
\newcommand*{\reporthead}{}% Report number

@@

The limitations of the approach here proposed can be appraised by examining
the assumptions and approximations on which it rests: three assumptions and
two approximation all in all. These become in turn clear within a
full-fledged probabilistic approach. We summarize them here and give a more
mathematical sketch in appendix~\ref{sec:inference_prob}.

The first and most important assumption of this kind of maximum-entropy
applications in neuroscience can be expressed in two equivalent ways:
\begin{enumerate*}[(\roman*)]
  % \item the order of the time bins has no relevance;
\item the frequencies of the activity during a recording are
  informationally sufficient for inferences about unrecorded times; \item
  our degree of belief about the sequence of activities is invariant under
  permutations of the time bins.
\end{enumerate*}
Equating time averages and expectations, \eqn~\eqref{eq:final_constraint},
would not make sense without this assumption. Its mathematical consequence
is to relate the probability for the frequency distribution $\yF$ during
the recording to that of the \emph{long-run} frequency distribution for the
activities of the larger population. Note that the maximum-entropy method
in statistical physics is meant to be used not with this assumption, but
with averages across \emph{repetitions} of the same experiment, because its
constraints represent macroscopically \emph{reproducible} quantities and
kinematics, even in non-equilibrium
\citep{jaynes1957,guntonetal1983,tikochinskyetal1984,grandy1988,berryetal1988,deroecketal2005_r2006,touchette2009}.

The second assumption of the maximum-entropy method concerns the form of
the probability distribution for the long-run frequencies ensuing from the
first assumption. It is taken to be either an \enquote{entropic prior}
\citep{skillingetal1984,rodriguez1991,neumann2007}, which takes into
account the multiplicity of long-run frequencies through a coefficient $L$,
or a prior for a model with sufficient statistics, which expresses the
sufficiency of the moments under study \citep{portamana2017}. This
assumption is crucial for obtaining the maximum-entropy method with the
Shannon entropy. A different density leads to alternative maximum-entropy
methods, for example based on the Burg entropy \citep{burg1975} if a
Dirichlet density is used \citep{jaynes1986d_r1996}.

The third assumption is specific to our application: all subsets of $\yA$
neurons in the larger population are equally likely to give rise to total
activity $\yA$, at every time bin. A mathematically equivalent assumption
is that $n$ neurons are sampled anew at each time bin. This assumption
allows us to factorize the joint probability of the two sequences of
activities and apply the sampling
relation~\eqref{eq:conditional_hypergeometric} of
appendix~\ref{sec:derivation_maxent_fullnet} at each bin.

Finally, the maximum-entropy method typically approximates the number $T$
of time bins, the prior coefficient $L$, and the ratio $T/L$ to infinity.
These approximations lead to the identification of the four distributions
listed at the end of \sect~\ref{sec:model}: the recorded frequency
distribution becomes equal to the long-run one, and also to the probability
distribution for the activity at recorded and new time bins. In our
application it also leads to the equality of the hypergeometric
distribution~\eqref{eq:conditional_hypergeometric} and the conditional
frequencies of the sample activity $\ya$ given $\yA$. The goodness of this
approximation decreases as the ratio between the number of possible joint
states and the number of bins -- roughly $nN/T$ -- increases, and also as
observed quantities approach their mathematical bounds, such as some
frequencies $\yff_{\ya}=0$. Without these approximations it is still
possible to interpret the maximum-entropy
distribution~\eqref{eq:app_maxent_pop_first} as the mode of the probability
distribution for the recorded frequencies $\yF$, that is, the most probable
recorded frequency.

The first and third assumptions above are the least realistic, but they
make sense as reference hypotheses for judging the informational relevance
of correlations of some order, as in \sect~\ref{sec:marginalization}, to be
compared with hypotheses about more realistic time dependences and
clustering.

The two infinity approximations are at the limit of their validity in our
specific application, so the distributions discussed in our example are
best interpreted as \emph{most probable frequency distributions}.

The space of possible frequency distributions for the larger population has
$N$ dimensions, however, and probability distributions in high dimensions
have counter-intuitive properties. For example, the mode or mean can have
\emph{atypical} features when compared with the features of most other
points of the space. The question, then, is \emph{which features of the
  maximum-entropy frequency distribution are typical of the majority of
  plausible frequency distributions?} We can only answer for sure by using
the full-fledged probability calculus. A more complete study (in
preparation) with the first data set reveals that most of the plausible
frequency distributions have three important features in common with the
maximum-entropy one:
\begin{itemize}
\item all activities $\yA/N \gtrsim 5\%$ have practically zero frequencies;
\item there are two regions of activity levels with high frequencies,
  roughly separated by a trough of lower frequencies.
\item The frequencies of the region on the left ($\yA/N \lesssim 1.8\%$)
  are higher than those of the region on the right ($\yA/N \gtrsim 1.8\%$).
\end{itemize}
But there are also differences. For example, many plausible frequency
distributions have three or four modes instead of just two; these modes are
higher than those of the maximum-entropy distribution; and the bump of high
frequencies on the right is slightly shifted towards lower activities than
the corresponding maximum in the maximum-entropy distribution.

@@

\section{Sketch of inference with the probability calculus}
\label{sec:inference_prob}
